Relict deciduous oaks show high resilience to drought despite the effect of land-use legacies along a climatic gradient in the rear-edge

### 2019-07-01 man/ecosystems/ms_ecosystems.Rmd
update conclusions

## Resilience to drought events at the stand and individual tree levels
The exploration of the resilience metrics of tree-growth for severe drought events since 1950 (*i.e.* for the share period among the three chronologies, excluding juvenile years, Table S3) revealed a positive relation between drought severity and Recovery (*Rc*), significant for all oak populations (Figure 6). A similar pattern was obtained for Resilience (*Rs*) but only significant for SJ site (Figure 6). Non-significant patterns were obtained if we exclude 1995 from the analysis (Figure S8). Trees showed the highest value of tree-growth resilience for 1995, the worst drought event in our study area (Table S3), and particularly SJ, where our results suggest a major release event after 1995 (Figure 5).

@@

# Conclussions 
`$TODO$` las conclusiones

Severe drought events have provoked important reductions in primary and secondary growth of *Q. pyrenaica* forests. However, despite the expected vulnerability for those relict stands, we observed positive trends in primary growth (since 2000) and secondary growth (since 1970) of *Q. pyrenaica* in Sierra Nevada mountain range. Surprisingly we observed high values of resilience to drought, particularly at long-term scales. These results are consistent with research showing high genetic resilience values and overall suggest this mountain region still acting as refuge for deciduous *Quercus*. The differences observed in tree-growth and resilience to drought between very close sites show that responses to drought were site-dependent. This is particularly relevant for rear-edge populations where topographic and biophysical variablity facilitates the existence of microrefugia [@VilaCabrera2019].

The analysis of tree-growth dynamics revealed suppression and release events that are consistent with socieconomic dynamic, derived from the review of historical documents. Recently works have combined information from both tree-ring and remote-sensing to analyze the resilience of forest to drought [*e.g.* @Gazol2018]. The approach presented here, which also includes an exhaustive review of historical documents, help to identify the response of vegetation to climate and to land use changes. In this sense our results highligth the importance of landuse legacies for highly transformed Mediterranean systemas [@Navarro2013].

Growth projections have forecasted a decrease in productivity for *Q. pyrenaica* that would increase vulnerability of this species to climate warming at the dry edges locations [@GeaIzquierdo2013]. However, the positive trends for both primary (*i.e.* greenness) and secondary growth (BAI) observed ar our study site and the high resilience values to drought showed at long term scale, seem to suggest that Sierra Nevada constitutes still a refuge for some species. This is relevant for tree species with higher sensitivity to climate change, such us *Q. pyrenaica* (García-Valdes et al. 2013; Benito-Garzón et al. 2008), not only for conservation per se of this species, but for all ecosystem services that these singular forests offer at Mediterranean mountain.

The abandonment of traditional activities along with the increase in temperature since the 70s, could be explaining the positive growth trend observed. This trend and the increase of tree-cover observed in the last decades in this mountain region [@CamachoOlmedo2002], would enhance the carbon sequestration of this forest promoting its role as carbon sink. However, the abandonment of traditional management has provoked a general state of degradation of this forest formation [@Canellas2004; @Corcuera2006]. In this sense, es necesario analizar con mas detalle como se verán afectados los servicios ecosistémicos que ofrecen estos robledales en su rear edge en un contexto de cambio global, identificando claramente los impactos del clima y del land-use change en la provisión de servicios ecosistémicos.

>>>>

## Resilience to drought events at the ecosystem and individual tree levels
The exploration of the resilience metrics of tree-growth for severe drought events since 1950 (*i.e.* for the share period among the three chronologies, excluding juvenile years, Table S3) revealed a positive relation between drought severity and Recovery (*Rc*), significant for all oak populations (Figure 6). A similar pattern was obtained for Resilience (*Rs*) but only significant for SJ site (Figure 6). Non-significant patterns were obtained if we exclude 1995 from the analysis (Figure S8). Trees showed the highest value of tree-growth resilience for 1995, the worst drought event in our study area (Table S3), and particularly SJ, where our results suggest a major release event after 1995 (Figure 5).

@@

# Conclussions 
Severe drought events have provoked important reductions in primary and secondary growth of *Q. pyrenaica* forests at Sierra Nevada mountain range. However, despite the expected vulnerability for those relict stands, we observed positive trends of both primary (*i.e.* greenness) at ecosystem scale and secondary growth at tree level in our study site, particularly at the most humid site along the climatic gradient analyzed. Surprisingly we observed high resilience values to drought, particularly at long-term scales. These results are consistent with research showing high genetic resilience values and overall suggest this mountain region still acting as refuge for deciduous *Quercus*.

The differences observed in tree-growth and resilience to drought between very close sites show that responses to drought were site-dependent. This is particularly relevant for rear-edge populations where topographic and biophysical variablity facilitates the existence of microrefugia, as reflected along the climatic gradient analysed [@VilaCabrera2019].

The analysis of tree-growth dynamics revealed suppression and release events that are consistent with legacies left by land-use in local forest dynamics, as inferred from an exhaustive review of historical documents. In this sense our results highlight the importance of land-use legacies for highly transformed Mediterranean systems. This is relevant for tree species with higher sensitivity to climate change, such us *Q. pyrenaica*, not only for conservation per se of this species, but for all ecosystem services that these singular forests offer at Mediterranean mountain.

Following our study it is necessary to disentangle how ecosystem services and stand dynamics are affected by the combination of climate change and land-use legacies in highly transformed landscapes like the one studied here.
